---
title: "My Project"
---


```{r}
#| label: load-packages
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| include: true

library(md)
library(tidyverse)
library(DBI)
library(duckdb)


```

## Overview

This is a collection of utilities to help with the management, administration and navigation of [duckdb](https://duckdb.org/) database either locally on your computer or in the cloud via [motherduck](https://motherduck.com/)

Database management is incredibly easy in R with fantastic packages such as [DBI](https://dbplyr.tidyverse.org/) and [dbplyr](https://dbplyr.tidyverse.org/), however some databases have specific extensions or utilities that are aren't readily accessible via this packages

{md} pack simplifies these common database administration task with easy to understand syntax. {md} is built upon [DBI](https://dbplyr.tidyverse.org/) and returns a lazy DBI object so that you can further fully integrate your data with [dbplyr](https://dbplyr.tidyverse.org/)


## Future ambition

Eventually, I'll use the learning from this package to create a meta DB utilities package so that regardless if you're in snowflake, DuckDB, Redshift, etc you will have generalized functions that work across your database types

This is very much work in progress -- I'll eventually transition to the [R7](https://rconsortium.github.io/S7/articles/S7.html) object system but just want to get some usage first before deciding on the architecture and structure.

Please create an issue if you have any comments or requests or reach out if you have any feedback.

## Overview of functions

Below is a quick overview of the functions available in the package.

### Functions that help you manage your connection and duckdb specific extensions

-   `connect_to_motherduck()` will leverage your motherduck token to connect you to your motherduck instance (it will install the motherduck extension if not already present)
-   `install_extensions()` will install various duckdb extensions from the official repository
-   `load_extensions()` will load a duckdb extensions either from an official repository 
-   `validate_md_connection_status()` will validate your motherduck connection status
-   `validate_extension_install_status()` will validate if an extension has been successfully installed
-   `validate_extension_load_status()` will validate if an extension has been successfully loaded

### Functions that help you see what is in your databases
-   `pwd()` prints the current database that you are "in" 
-   `cd()` will change your "root" database so any execution functions are relative to that database (eg. CREATE SCHEMA)
-   `list_database()` list the databases and their metadata
-   `list_schema()`list the schemas and their metadata
-   `list_table()`list the tables and their metadata
-   `list_all_table()`list all tables across all databases
-   `list_view()`list the views and their metadata

### Functions that will help you read data into duckdb or motherduck 
-   `read_httpfs()` will read httpfs file formats 
-   `read_parquet()` will read parquet file formats
-   `read_excel()` will read excel files formats 

### Functions that will help you create or replace databases, scehems, tables or views 
-   `create_or_replace_database()` will take R data and create a database with your data
-   `create_or_replace_schemas()` will take R data and create a schema with your data
-   `create_or_replace_view()` will take R data and create a views with your data
-   `create_or_replace_table()` will take R data and create a tables with your data
-   `drop_table()` will delete a table from your databases
-   `drop_database()` will delete a database


## Functions to help you manage motherduck users, tokens, and instance settings
-   `list_md_active_accounts()` list users with active duckling instances (note: will not list inactive users)
-   `list_md_user_instance()` list a user's instance settings
-   `list_md_user_tokens()` list user's tokens
-   `show_current_user()` show your current user name
-   `create_md_user()` create user or service account in your organization
-   `delete_md_user()` delete a user or service account
-   `create_md_access_token()` create an access token for a user
-   `delete_md_access_token()` delete an access token for a user
-   `configure_md_user_settings()` configure a user's instance settings

### Functions to help you understand your data
-   `summary()` will summarize your table or view's data

## What do I need to use this?

-   [duckdb](https://duckdb.org/) R package installed on your computer
-   A [motherduck](https://motherduck.com/) account
-   A motherduck access token which you you can be saved to your R environment file with `usethis::edit_r_environ()`



```{r}
#| label: connection-example
#| echo: true
#| eval: true
#| warning: true
#| message: true
#| include: true


con_md <- connect_to_motherduck("MOTHERDUCK_TOKEN")

```

